Taking uncertainty in word embedding bias estimation seriously: a  bayesian

### 2021-12-10 paperDraft/paperDraft2.Rmd
setting up paper draft 2

---
title: 'Taking uncertainty in word embedding bias estimation seriously: a  bayesian
  approach'
author: "Alicja Dobrzeniecka and Rafal Urbaniak"
output:
  pdf_document:
    number_sections: yes
    df_print: kable
    keep_tex: yes
    includes:
      in_header: Rafal_latex6.sty
  html_document:
    df_print: paged
classoption: dvipsnames,enabledeprecatedfontcommands
fontsize: 10pt
documentclass: scrartcl
urlcolor: blue
bibliography: ../references/bibliographyRafal.bib
csl: ../references/apa-6th-edition.csl
---

@@

\tableofcontents

@@

@@

A similar methodology is employed by @Garg2018years, who employ word embeddings trained on corpora from different decades to study the shifts in various biases. For instance, to compute the occupational embeddings bias for women the authors first compute the average vector of vector emeddings of words that represent women (e.g. "she", "female"), then calculate the Euclidean distance between this mean vector and words for occupations. Then they take the mean of these distances and subtract from it the analogously obtained mean for the average vector of vector embeddings of words that represent men. Formally they take the relative norm distance between $X$ and $Y$ to be:
\begin{align}
\textsf{relative norm distance} & = \sum_{v_m\in M} \vert \vert v_m - v_X\vert \vert_2 - \vert v_m - v_Y\vert \vert_2
\end{align}
\noindent where the norm used is Euclidean, and $v_X$ and $x_Y$ are average vectors for sets $X$ and $Y$ respectively.

@@

@@

Such approaches, however, are not unproblematic. One may cosinder the selection of attributes for the bias research. For one thing, the word lists are often fairly small, and the papers do not discuss the impact of word list sizes on the uncertainty involved. 
We will argue that the word list sizes are too small to yield strong conclusions, given the variances involved.

@@

# The problem with means of means

<!-- A <- c(1,5,10) -->
<!-- B <- c(6, 8, 15) -->

<!-- c(A,B) -->

<!-- sd(c(A,B)) -->

<!-- mA <- mean(A) -->
<!-- mB <- mean(B) -->

<!-- sd(c(mA,mB)) -->

The measures described all calculate means of means and their authors run statistical tests on sets of means. This, however, is problematic for two related reasons. One, by pre-averaging data we throw away information about sample sizes. For the former point, think about proportions: 10 out of 20 and 2 out of 4 give the same mean, but you would obtain more information by making the former observation rather than by making the latter.  Two, when we pre-average, we remove variation, and so pre-averaging  tend to manufacture false confidence. We will have more to say about whether this happens in the case of applications of MAC, for now let's go over a simple example to make the conceptual point clear.

To illustrate let's employ the formulas used by @Caliskan2017semanticsBiases in a simple example. Conceptually, all such tests come up with rather short lists of protected words and rather short lists of stereotypical attributes. Clearly, these are not complete list. So let's treat them as  samples from richer pools of stereotypical predicates and let's take the uncertainty and variation involved seriously.

Consider a simple situation in which there are two protected classes, $X=\{t_1,t_2\}$ and $Y=\{t_3,t_4\}$ and two five-element attribute sets $A$ and $B$.

First, we play around with a scenario in which all the protected terms are on average equally unsimilar to both sets ($\mu =0$) with standard deviation of $.05$. Let's randomly draw similarity scores and plot the results with group means plotted as vertical lines.

\footnotesize
```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
set.seed(123)
t1 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
t2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
t3 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
t4 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
```
\normalsize

```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
xlimits <- c(-.3,.3)

t1Plot <- ggplot(t1)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t1 in X")+xlim(xlimits)

t2Plot <- ggplot(t2)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t2 in X")+xlim(xlimits)

t3Plot <- ggplot(t3)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t3 in Y")+xlim(xlimits)

t4Plot <- ggplot(t4)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t4 in Y")+xlim(xlimits)

grid.arrange(t1Plot,t2Plot, t3Plot, t4Plot, ncol=2)
```

\noindent When you look at the datapoints, do you have the impression of a strong bias here? We wouldn't think so.  But now let's run the calculations from [@Caliskan2017semanticsBiases].

\vspace{1mm}
\footnotesize
```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
s <- function (table){ mean(table$A) - mean(table$B)}
whole <- rbind(t1,t2,t3,t4)
rawsd <- sd(c(whole$A,whole$B))
factor <- sd(c(s(t1),s(t2),s(t3),s(t4)))
rawsd/factor
numerator <-  mean(s(t1),s(t2)) - mean(s(t3),s(t4))
print(list(factor = factor,numerator = numerator, bias = numerator / factor))
```
\normalsize

\noindent This should make us pause. We know these were points randomly drawn from distributions where there is no difference in means. Yet, the calculated effect size is 1.82, whereas the largest effect size reported by @Caliskan2017semanticsBiases is 1.81!

Interestingly, if we repeat the drawing 10000 times, each time calculating the  bias, it turns out that with this variance and sample size, pretty much anything can happen.

\vspace{1mm}
\footnotesize
```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}

biasesNull <- numeric(10000)
for(i in 1:10000){
t1 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
t2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
t3 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
t4 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))

factor <- sd(c(s(t1),s(t2),s(t3),s(t4)))
numerator <-  mean(s(t1),s(t2)) - mean(s(t3),s(t4))
biasesNull[i]  <- numerator / factor
}
ggplot()+geom_histogram(aes(x=biasesNull, y = ..density..), alpha = 0.6, bins=50)+
  theme_tufte()+labs(title="10k biases for identical means and sd =.05")+ xlab("bias")
```
\normalsize

Now, let's simulate a situation where the means are identical but the standard deviation is much smaller, .001.

\footnotesize
```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
set.seed(124)
t1v <- data.frame(A  = rnorm(5,0,0.001), B = rnorm(5,0,0.001))
t2v <- data.frame(A  = rnorm(5,0,0.001), B = rnorm(5,0,0.001))
t3v <- data.frame(A  = rnorm(5,0,0.001), B = rnorm(5,0,0.001))
t4v <- data.frame(A  = rnorm(5,0,0.001), B = rnorm(5,0,0.001))
```
\normalsize

```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
xlimits <- c(-.3,.3)

t1vPlot <- ggplot(t1v)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t1 in X")+xlim(xlimits)

t2vPlot <- ggplot(t2v)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t2 in X")+xlim(xlimits)

t3vPlot <- ggplot(t3v)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t3 in Y")+xlim(xlimits)

t4vPlot <- ggplot(t4v)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t4 in Y")+xlim(xlimits)

grid.arrange(t1vPlot,t2vPlot, t3vPlot, t4vPlot, ncol=2)
```

\noindent When you look at the datapoints, do you have the impression of a strong bias here? We wouldn't think so.  But now let's run the calculations from [@Caliskan2017semanticsBiases].

\vspace{1mm}
\footnotesize
```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
factorV <- sd(c(s(t1v),s(t2v),s(t3v),s(t4v)))
numeratorV <-  mean(s(t1v),s(t2v)) - mean(s(t3v),s(t4v))
print(list(factor = factorV,numerator = numeratorV, bias = numeratorV / factorV))
```
\normalsize

While the numerator and the factors changed a lot, the bias actually increased. One reason bias increases is that once the standard deviation goes down, so does the factor used in the calculation of bias.

Again, to see whether this metric would provide us with meaningful information, let's simulate 10000 drawings.

\vspace{1mm}
\footnotesize
```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
set.seed(124)

biasesLowVariance <- numeric(10000)
for(i in 1:10000){
t1v <- data.frame(A  = rnorm(5,0,0.001), B = rnorm(5,0,0.001))
t2v <- data.frame(A  = rnorm(5,0,0.001), B = rnorm(5,0,0.001))
t3v <- data.frame(A  = rnorm(5,0,0.001), B = rnorm(5,0,0.001))
t4v <- data.frame(A  = rnorm(5,0,0.001), B = rnorm(5,0,0.001))

factorV <- sd(c(s(t1v),s(t2v),s(t3v),s(t4v)))

numeratorV <-  mean(s(t1v),s(t2v)) - mean(s(t3v),s(t4v))

biasesLowVariance[i] <- numeratorV / factorV
}
ggplot()+geom_histogram(aes(x=biasesLowVariance, y = ..density..), alpha = 0.6, bins=50)+
  theme_tufte()+labs(title="10k biases for identical means and sd =.001")+ xlab("bias")
```
\normalsize

Again, not so informative. Now, let's try sampling from distributions where there in fact is a difference in means. Terms from $X$ are on average .1 similar to $A$ (and still $0$ to $B$), while terms from $Y$ are $.1$ similar to $B$ (and 0 to $A$). The standard deviation is 0.05 in all the cases. There is a  clear difference between $X$ and $Y$ and quick visual inspection should tell us so.

\vspace{1mm}
\footnotesize
```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=FALSE, fig.show = "hold", out.width = "100%"}
set.seed(766)
t1d2 <- data.frame(A  = rnorm(5,.1,0.05), B = rnorm(5,0,0.05))
t2d2 <- data.frame(A  = rnorm(5,.1,0.05), B = rnorm(5,0,0.05))
t3d2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,.1,0.05))
t4d2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,.1,0.05))
```
\normalsize

\vspace{1mm}
\footnotesize
```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
library(grid)
library(gridExtra)
t1d2Plot <- ggplot(t1d2)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t1 in X")+xlim(xlimits)

t2d2Plot <- ggplot(t2d2)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t2 in X")+xlim(xlimits)

t3d2Plot <- ggplot(t3d2)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t3 in Y")+xlim(xlimits)

t4d2Plot <- ggplot(t4d2)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t4 in Y")+xlim(xlimits)

grid.arrange(t1d2Plot,t2d2Plot, t3d2Plot, t4d2Plot, ncol=2,   
      top = textGrob("different means, sd=0.05",gp=gpar(fontsize=15,font=1)))
```
\normalsize

\noindent  Is this clear difference mirrored in the calculations?

\footnotesize
```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=FALSE, fig.show = "hold", out.width = "100%"}
factorD2 <- sd(c(s(t1d2),s(t2d2),s(t3d2),s(t4d2)))
numeratorD2 <-  mean(s(t1d2),s(t2d2)) - mean(s(t3d2),s(t4d2))
biasD2 <- numeratorD2 / factorD2
biasD2
```
\normalsize

\noindent The absolute value of the effect size is smaller than in the null case with the same standard deviation.  Let's simulate 10000 drawings:

\vspace{1mm}
\footnotesize
```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
biasesD2 <- numeric(10000)
for(i in 1:10000){
t1d2 <- data.frame(A  = rnorm(5,.1,0.05), B = rnorm(5,0,0.05))
t2d2 <- data.frame(A  = rnorm(5,.1,0.05), B = rnorm(5,0,0.05))
t3d2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,.1,0.05))
t4d2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,.1,0.05))

factorD2 <- sd(c(s(t1d2),s(t2d2),s(t3d2),s(t4d2)))
numeratorD2 <-  mean(s(t1d2),s(t2d2)) - mean(s(t3d2),s(t4d2))
biasesD2[i] <- numeratorD2/factorD2
}

ggplot()+geom_histogram(aes(x=biasesD2, y = ..density..), alpha = 0.6, bins=50)+
  theme_tufte()+labs(title="10k biases for different means and sd =.05")+ xlab("bias")
```
\normalsize

\noindent Now suppose we keep the means the same but increase the standard deviation to .1.

\vspace{1mm}
\footnotesize
```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=FALSE, fig.show = "hold", out.width = "100%"}
set.seed(766)
t1d1 <- data.frame(A  = rnorm(5,.1,0.1), B = rnorm(5,0,0.1))
t2d1 <- data.frame(A  = rnorm(5,.1,0.1), B = rnorm(5,0,0.1))
t3d1 <- data.frame(A  = rnorm(5,0,0.1), B = rnorm(5,.1,0.1))
t4d1 <- data.frame(A  = rnorm(5,0,0.1), B = rnorm(5,.1,0.1))
```
\normalsize

\vspace{1mm}
\footnotesize
```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
t1d1Plot <- ggplot(t1d1)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t1 in X")+xlim(xlimits)

t2d1Plot <- ggplot(t2d1)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t2 in X")+xlim(xlimits)

t3d1Plot <- ggplot(t3d1)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t3 in Y")+xlim(xlimits)

t4d1Plot <- ggplot(t4d1)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t4 in Y")+xlim(xlimits)

grid.arrange(t1d1Plot,t2d1Plot, t3d1Plot, t4d1Plot, ncol=2)
```
\normalsize

\noindent  Is this clear difference mirrored in the calculations?

\footnotesize
```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=FALSE, fig.show = "hold", out.width = "100%"}
factorD1 <- sd(c(s(t1d1),s(t2d1),s(t3d1),s(t4d1)))
numeratorD1 <-  mean(s(t1d1),s(t2d1)) - mean(s(t3d1),s(t4d1))
biasD1 <- numeratorD1 / factorD1
biasD1
```
\normalsize

\noindent The absolute value of the effect size is smaller than in the previous case!

\vspace{1mm}
\footnotesize
```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
biasesD1 <- numeric(10000)

for(i in 1:10000){
t1d1 <- data.frame(A  = rnorm(5,.1,0.1), B = rnorm(5,0,0.1))
t2d1 <- data.frame(A  = rnorm(5,.1,0.1), B = rnorm(5,0,0.1))
t3d1 <- data.frame(A  = rnorm(5,0,0.1), B = rnorm(5,.1,0.1))
t4d1 <- data.frame(A  = rnorm(5,0,0.1), B = rnorm(5,.1,0.1))

factorD1 <- sd(c(s(t1d1),s(t2d1),s(t3d1),s(t4d1)))
numeratorD1 <-  mean(s(t1d1),s(t2d1)) - mean(s(t3d1),s(t4d1))
biasesD1[i] <- numeratorD1/factorD1
}

ggplot()+geom_histogram(aes(x=biasesD1, y = ..density..), alpha = 0.6, bins=50)+
  theme_tufte()+labs(title="10k biases for different means and sd =.001")+ xlab("bias")
```
\normalsize

This is is a bit better, but still quite some uncertainty is involved, far from what systematically low mean-based p-values reported in the papers might suggest.  Of course, this is a bit of a caricature, as our word lists were short (four protected words and 10 attributes). But the word lists used in the actually papers are not much longer.  In such a set-up the key observations are as follows:

- Seemingly high bias measures might arise even if the underlying processes actually have the same mean.

- Even if the mean remains the same, non-negligible changes can result from a shift in the standard deviation of the original process, and the change might go in the opposite direction than a visualisation of datapoints might suggest, because with the decrease of standard deviation, the factor decreases and the bias increases.

- Even if the underlying means are the same, but the variation is different, the bias metric in the long run could tend toward a different value:

- The lack of control group in the paper and our analysis indicates that without neutral baseline it is difficult
to interpret the effectiveness of the metric.

\vspace{1mm}
\footnotesize
```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
mean(biasesD1); mean(biasesD2)
median(biasesD1); median(biasesD2)
```

\normalsize
and so the point estimations of bias are sensitive to factors other than the underlying process means.

- Even if there is a difference in means, the bias metric can be lower, and the uncertainty about it needs to be gauged.

- The uncertainty resulting from including the raw datapoint variance into considerations is more extensive than the one suggested by the low p-values obtained from taking means as datapoints.

\vspace{1mm}
\footnotesize
```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
quantile(biasesD2, probs = c(0.275,0.975))
quantile(biasesD1, probs = c(0.275,0.975))
```
\normalsize

# Bayesian estimation

One thing to notice is that in @Manzini2019blackToCriminal words from all three subclasses (in the case of religion by a subclass we mean Christianity, Judaism, and Islam) were compared against all of the stereotypes, which means that there was no distinction between cases in which the stereotype is associated with a given sub-class, as opposed to the situation in which it is associated with another one. Not all of the stereotypical words have to be considered as harmful for all of the sub-groups. One should investigate the sub-groups separately as some of them may have stronger harmful associations than others.

We decided to add control groups in the form of two classes — neutral words and human-related words. Without a proper control group it is quite hard to compare the resulting cosine distances and decide on their significance in bias detection. We prepared approximately 230 neutral words to double-check the prima-facie neutral hypothesis that their cosine similarity to the protected words will oscillate around 0 (that is, the distances will be around 1). This provides us with a more reliable point of reference. Moreover,
we added human attributes that are associated with people in general to investigate whether the smaller cosine distance between protected words and stereotypes can result simply from the fact that the stereotype predicates are associated with humans.

One may consider improving the bias detection method presented in @Manzini2019blackToCriminal. As the chosen protected words and attributes do not represent the whole biased population but only a sample, it is justified to introduce the uncertainty measure to the method. We decided to use Bayesian estimation to make the cosine distance method more methodologically correct.

Our prior distributions for the Bayesian model reflect our first stab at what we know or expect before we see the data.  All our distributions are fairly weak. The distribution for mean distance (cosine distance) is set to fit (almost completely) between 0 and 2 with the mean equal to 1, in line with the idea that  most of the cosine distances may be considered neutral and therefore their similarity will be around 0. That is, we take it to be DISTRIBUTION. The  prior distribution for standard deviation is half-cauchy between 0 and 1, but unlikely to be as large as 1. Finally the prior distributions for  the coefficients have most of their mass between -1 and 1, with mean around 0.  As we want to put ourselves in skeptical shoes looking at   @Manzini2019blackToCriminal and asking whether the data would indeed convince us of the existence of bias, we decided to use slightly skeptical (mean 0), but  fairly weak (sd = ) as our regularizing priors. The model parameters will have a posterior distribution obtained using either Hamiltionian Monte Carlo methods (STAN) available through the rethinking package.

\begin{center}
\begin{figure}[!htb]\centering
   \begin{minipage}{0.55\textwidth}
  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
   \end{minipage}
   \begin {minipage}{0.43\textwidth}
    \includegraphics[width=7cm]{../images/religionGoogleCoeffs.jpeg}
   \end{minipage}

  \begin{minipage}{0.55\textwidth}
\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
\end{minipage}
   \begin {minipage}{0.43\textwidth}
    \includegraphics[width=7cm]{../images/genderGoogleCoeffs.jpeg}
   \end{minipage}

  \begin{minipage}{0.55\textwidth}
\includegraphics[width=7cm]{../images/raceCoeffs.jpeg}
\end{minipage}
   \begin {minipage}{0.43\textwidth}
    \includegraphics[width=7cm]{../images/raceGoogleCoeffs.jpeg}
   \end{minipage}
\end{figure}

\end{center}

Let's analyse dataset-level coefficients together with their highest posterior density intervals, for the three topic groups. We compare the results for two different word emebeddings corpuses - RedditL2 and GoogleNews. For Google embeddings the HPDIs
for all class coefficients (associated, different, human, and none) include zero. This can lead one to a conclusion that the impact for associated, different, human and neutral attribute is, when averaged, quite similar. This indicates how including the uncertainty
may change the use and interpretation of Manzini, Lim, Tsvetkov, & Black (2019) MAC metric. It seems that if one focuses only on differences between means of means, this is too simplistic.

In the case of Reddit word embeddings the situation is similar although HPDI is below 0 for Gender class when looking at `associated` and `different` mean coefficient. This can suggest that there is indeed slightly stronger impact of these attributes. One should also notice how in general associated and different coefficients have quite similar HPDI range, the highest observed absolute difference is equal to only 0.1. This suggests that again the impact of associated attributes and different ones is not clear at first sight. Finally, let’s compare the HPDIs for Reddit and Reddit debiased datasets. For Religion and Race dataset there is a minor shift (in absolute values the highest change is equal to approximately 0.1) of the mean coefficients towards zero. However for Gender dataset there is no significant change. This is of course a general look at the group coefficients. Let’s now analyze the individual words.

Looking at Reddit individual words one can observe that for Religion the cosine distance results for the associated attributes are for most of the words close to human attributes as well. This can suggest that in some cases words concerning humans can have higher
similarity with some protected words independently of whether they are stereotypical or neutral-human words. One should also pay attention to the fact that for all of the associated and different attributes in Religion, the uncertainty intervals overlap at
some point. What is even more surprising is that for protected words, such as “torah” the associated attribute has the the cosine distance slightly over 1, which means no positive similarity! If the protected words that we chose do not have high similarity
with harmful associated attributes, then one should consider at least three scenarios. The first one is that the choice of the protected words and attributes may be corrupted. The second one is that the metric is unable to catch the hidden bias properly. The third one is that there is actually no bias between the words. Regardless of which scenario one considers, it is essential to take a look at the individual values before averaging them or aggregating in other ways. It seems that using Bayesian method can enhance
the process of verifying the hypotheses concerning the choice of protected words and attributes.

Surprisingly in Gender one can observe high cosine similarity values between some female stereotypical professions and male protected words. If a word stereotypically associated with females has low cosine distance to male protected words, then one should investigate the issue further. The reasons for this unexpected cosine distance may lie in the frequency of appearance of the protected words in the raw data. Some of the groups (like Muslim people or females) may have less representation in the data that is taken as an input for word embeddings. Therefore, if we assume that the MAC detects actually co-occurrence only, it makes sense that they can have high similarity with associated attributes and lower similarity with different ones. At the same time male protected words and other religions can have high similarity with different attributes because they have greater representation in the dataset in general and occur close to much more concepts. Cosine distance seems to capture the information on the co-occurrence of words and not on the semantic similarity strictly speaking.

In Google, one may notice interesting phenomena as well. Although the GoogleNews dataset is larger, trained on different data sources and with the use of more dimensions, some of the results are similar to the ones obtained for the Reddit corpus. For Religion
almost all of the values for associated class intersect with different class as well. In the case of Race it is 100% of the available words. This indicates again that it is not clear how the metric should be used. Quite a different situation takes place for
Gender, where the similarity between associated and different is present mostly for the male protected words. This means that male words have high similarity with both male stereotypical professions and female ones. However, for females, the similarity is
high mostly only for the female stereotypical professions. This observation could not be made when using MAC metric only, as it requires investigating the individual words, and providing uncertainty.

# Effects of debiasing

One may verify if the dataset-level coefficients changed after the debiasing method used in @Manzini2019blackToCriminal. 
In some cases there is a minor change to the right, the mean value is closer to 0. The `associated` and `differen` attributes are now more overlapping than before. However, the uncertainty intervals are still high and therefore the significance of the change cannot be considered as enormous. The surprising obervation is that in the case of `Gender` and `Race` groups, the neutral words shifted to the left, further from the neutral position.

\begin{center}
\begin{figure}[!htb]\centering
   \begin{minipage}{0.55\textwidth}
  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
   \end{minipage}
   \begin {minipage}{0.43\textwidth}
    \includegraphics[width=7cm]{../images/debiasedReligionRedditCoeffs.jpeg}
   \end{minipage}

>>>>

---
title: 'Taking uncertainty in word embedding bias estimation seriously: a  bayesian
  approach'
author: "Alicja Dobrzeniecka and Rafal Urbaniak"
output:
  pdf_document:
    number_sections: yes
    df_print: kable
    keep_tex: yes
    includes:
      in_header: Rafal_latex7.sty
  html_document:
    df_print: paged
classoption: dvipsnames,enabledeprecatedfontcommands
fontsize: 10pt
documentclass: scrartcl
urlcolor: blue
bibliography: ../references/cosineReferences1.bib
csl: ../references/apa-6th-edition.csl
---

@@

@@

\raf{fix bibliography}

\ali{proofread and think if we should add more}

@@

A similar methodology is employed by @Garg2018years, who employ word embeddings trained on corpora from different decades to study the shifts in various biases. For instance, to compute the occupational embeddings bias for women the authors first compute the average vector of vector embeddings of words that represent women (e.g. "she", "female"), then calculate the Euclidean distance between this mean vector and words for occupations. Then they take the mean of these distances and subtract from it the analogously obtained mean for the average vector of vector embeddings of words that represent men. Formally they take the relative norm distance between $X$ and $Y$ to be:
\begin{align}
\textsf{relative norm distance} & = \sum_{v_m\in M} \vert \vert v_m - v_X\vert \vert_2 - \vert v_m - v_Y\vert \vert_2
\end{align}
\noindent where the norm used is Euclidean, and $v_X$ and $x_Y$ are average vectors for sets $X$ and $Y$ respectively.

@@

\todo{odleglosc nie musi uchwytywac relacji semantycznej zeby oceniac bias}

\todo{zalozenie, bardziej ze jezeli odleglosci przekladaja sie na downstream, to warto patrzec na odleglosci, nawet bez glebszej filozoficznej interpretacji ich}

@@

Such approaches, however, are not unproblematic. One may consider the selection of attributes for the bias research. For one thing, the word lists are often fairly small, and the papers do not discuss the impact of word list sizes on the uncertainty involved. 
We will argue that the word list sizes are too small to yield strong conclusions, given the variances involved.

@@

\todo{MAC: avoid looking at only one vector}

\todo{effect sizes}

# Library for MAC
